Depression and suicidality as costly signals of need

### 2021-04-12 Paper.rmd
Tweaked demographics plot and a couple of captions.

Major depression and suicide are generally viewed as pathological responses to stress that elicit aversive responses from others. Alternatively, depression and suicide attempts might be the result of adaptations for reliably signaling need and bargaining for increased support from reluctant others. To address this hypothesis, we examined responses to emotional signals in an experimental vignette study involving claims of substantial need in the presence of conflicts of interest and private information about whether the signaler was actually in need. In our sample of 1,240 participants recruited from Amazon Mechanical Turk, costlier signals like depression and suicide attempts resulted in greater perceptions of need and increased likelihood of support compared to simple verbal requests and crying without further symptoms. Additionally, the effect of signal type on likelihood of support was largely mediated by the effect of signaling on belief in need as predicted. Our results suggest that depression and suicide attempts, apparent human universals, are a relatively successful way of signaling honest need and eliciting support, consistent with costly signaling explanations.

@@

This strong connection between negative life events and depression along with the adversity and distress that likely accompanies these events has led to the suggestion that this relationship is causal [@KendlerEtAl1999]. Support for this comes from the fact that the relationship between negative events and depression holds true when looking solely at events outside of one's control, showing that the connection is unlikely to be driven solely by individuals who are already depressed creating situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. In addition to this, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity and that part of the heritability of depression stems from the heritability of negative events, like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999].

Of the types of negative life events associated with depression, those which involve conflict tend to be more strongly associated with depression [@Hammen2005; @Mazure1998]. Marital problems, bullying, and the existence of abusive relationships are all commonly cited risk factors for depression [@KendlerEtAl1995; @KendlerEtAl1999; @klomek2007; @klomek2019], with  assault (both sexual and non-sexual), in particular, greatly increasing one's risk of depression [@KendlerEtAl1995; @KendlerEtAl1999].  Other notable risk factors like loss of a loved one, financial problems, job loss, or severe or prolonged illness, might seem less related to conflict. However, the ability for others to provide meaningful support to limit the harm caused in times of need [@SugiyamaSugiyama2003] combined with the ubiquity of conflicts of interests mean that many types of problems will eventually involve social conflict even if they did not start that way [@Hagen2003]. In line with this reasoning, conflict, as a whole, was found to be associated with depression among the Tsimane, particularly when it involved non-kin [@StieglitzEtAl2015a].

@@

One possible reason for the connection between depression, suicide, and social adversity is that some instances of depression and suicide attempts serve as means to elicit increased support from others [@andrews2006; @Hagen2002; @Hagen2003; @HagenEtAl2008; @nock2008; @Rosenthal1993; @Stengel1956; @SymeEtAl2016], particularly in times when an individual cannot solve substantial fitness-relevant challenges on their own [@Hagen2003; @SymeEtAl2016]. According to the bargaining models of depression and suicide, the small groups humans lived in for much of our evolutionary history created many situations in which certain forms of cooperation were unlikely to be easily replaced by others while also allowing for harmful or unfair social relationships to develop [@Hagen2003; @SymeEtAl2016]. In these instances, one's monopoly on one or more types of cooperation could be leveraged to change how one is treated if others valued an individual's cooperation sufficiently and believed them to be in genuine need.

Due to the ease in which signals can be faked [@Zahavi1975; @mokkonen2016] and the costs of accepting a deceptive signal as reliable [@dawkins1978; @johnstone1993; @Hasson1994], demonstrating that one’s need is genuine has likely been a recurrent adaptive problem since humans evolved the capacity to provide instrumental support to one another. One possible solution to this is to produce signals that are unprofitable to fake due to the costs involved. [@Zahavi1975]. This is possible whenever signaling is comparatively cheaper for those who have the relevant underlying quality relative to those who lack it [@grafen1990]. In these instances, individuals with the quality maximize their fitness at greater levels of investment in the signal than those who lack the quality or possess it to a lesser degree [@zahavi1977]. The end result of this is expected to be a signal that either tends to not be worth faking or is impossible to fake to a large degree due to selection against fakers constraining the physiology behind the signal [@zahavi1977; @gangestad2005].

Depression and suicidal behavior fit this criterion as both are relatively costly for those not facing adversity than those who are. In the case of depression, the large-scale reduction in one's normal routine of fitness enhancing behavior is relatively less costly for those whose situations result in a limited ability to obtain fitness-relevant benefits, with depression likely adding no additional costs in instances of extreme deprivation [@Hagen2003]. When it comes to suicide, the bulk of the costs associated with suicide attempts are only felt in times that result in death or severe bodily harm. For this reason, the way others view and respond to suicide attempts is likely to be strongly influenced by the probability that the attempt would result in death or severe injury [@Rosenthal1993]. Despite this difference, the previously mentioned logic still applies with increased risk of death being relatively more costly for those able to advance their fitness prospects in their day-to-day lives.

The costs of depression and suicidal behavior may be particularly important in times of conflict or when the level of support needed to address one’s situation is substantial, both of which should be expected to increase the payoff of manipulating others through fake signals. When either condition applies, less costly signals like verbal requests, facial sadness, and crying are likely to be less reliable as their low costs can easily be made up for with increased social support, thereby favoring the use of costlier signals so long as the costs of signaling are outweighed by the benefits [@SymeHagen2021].

Beyond signaling need to those invested in the well-being of the signaler, depression and suicidal behavior may also help an individual bargain for increased support [@Hagen2003]. This is possible due to the reduced investment in others that frequently accompanies depression or the threat to future cooperation that comes from suicide. In these instances, depression and suicide attempts may function analogously to a labor strike, sending the message that one’s need is genuine and that if the target of the signal wanted to continue receiving benefits from the depressed, they would need to increase support towards the signaler. Due to this bargaining component, aversive responses from the targets of the signal to the signaler are expected throughout the bargaining process, with potential helpers likely signaling the costliness of providing support. Importantly, such responses do not prevent the delivery of meaningful benefits, they are simply expected to reduce the risk of helpers providing too much support for too little need.

<!-- reword this paragraph -->

@@

Distinguishing between the information provided by signals versus cues is a major challenge for signaling approaches to behavior. A signal differs from a cue in that it is intentionally sent by an organism to alter the perceptions and behaviors of others, while a cue is any piece of information that influences an individual's behavior or physiological state that was not created for communicative effect [@MaynardSmithHarper1995]. Regarding the signals of interest, suicide attempts are more likely to take the form of pure, external signals than other symptoms of depression. This is because low mood's impact on behavior for non-signaling benefits likely conveys information as cues in addition to what might exist for signaling purposes. In other words, the signaling component of depression may come from the upregulation of symptoms beyond what would be optimal for non-signaling purposes rather than the creation of entirely new symptoms. Suicidality is likely an exception to this, due to the likelihood most suicide attempts likely do not have non-signaling functions [see @deCatanzaro1991 for a possible exception].

@@

For depression and suicide attempts to be worth the costs involved with signal creation, meaningful social support would be required [@Hagen2003; @SymeEtAl2016]. Most studies examining responses to depression have been observational and focused on responses outside of changes in social support. In general, most report aversive responses and high levels of rejection towards the depressed, and depressed individuals are often linked to impaired social functioning [@EvraireDozois2011; @GotlibLee1989; @HirschfeldEtAl2000]. One explanation for this is that depressed individuals engage in behaviors that drive others away. In line with this reasoning, both excessive reassurance seeking and negative feedback seeking, which often accompany depression, have been shown to elicit negative emotional responses from others and increase the chance of rejection from others [@JoinerMetalsky1995; @StarrDavila2008]. Negative interactions are also seen in families as depressed children tend to report less support and more negative responses from parents than non-depressed children, although it has not been demonstrated that such interactions did not lead to depression in the first place [@MesserGross1995; @SchwartzEtAl2012].

Despite being looked for less frequently, beneficial responses to depressed individuals are also reported. Depression has been associated with reduced likelihood of aggression within families [@hops1987; @dadds1992], including those already experiencing high levels of conflict [@SheeberEtAl2001]. There is also limited evidence that depression can result in increased levels of support from others outside the family. For example, non-depressed individuals were shown to be more cooperative and exhibit more caretaking behavior towards their college depressed roommates over the course of their time living together [@HokansonEtAl1986]. Likewise, a study examining responses to confederates trained to appear ‘depressed’, ‘anxious’, or ‘normal’ found that individuals were more likely to provide advice and support to ‘depressed’ individuals than ‘anxious’ or ‘normal’ individuals despite the confederate in the ‘depression’ condition being viewed more negatively than the other conditions and the ‘depression’ condition involving similar levels of distress compared to the ‘anxious’ condition [@stephens1987].

Most studies examining responses to those who have survived suicide attempts have focused on stigmatization. In these studies, perceptions of survivors as being weak, selfish, mentally ill, and antisocial are all being commonly reported [@BatterhamEtAl2013; @TzengLipson2004], with stigmatization often being found within and outside one's social network [@FreyEtAl2016; @ScoccoEtAl2012]. Despite this potential for stigmatization, increased social support and changes to important relationships have been reported to follow suicide attempts with some indication that these effects may hold long term [@Stengel1956]. For example, a study of 100 women who survived suicide attempts found that individuals gained identifiable benefits through the attempt in 75 cases, with 41 individuals benefiting from reconciliations with others [@Lukianowicz1971]. Unlike many Western countries, where suicide is often viewed as pathological [@Hidaka2012], members of traditional societies have been reported to view suicide attempts as cries for help rather than mental illness [@Shostak1981], with attempts having been described as ways of escaping unwanted marriage arrangements, persistent abuse, or a lack of support in obtaining mates by those engaging in the behavior and observers [@GutierrezdePinedaMuirden1948; @Hilger1957; @Karsten1935; @Tessmann1930; @Wilson1960]. Although findings that individuals view suicide attempts as cries for help is not necessarily evidence for their ability to lead to beneficial responses, 30 out of 84 examples of suicidal behavior included in the HRAF resulted in positive changes for the survivor, despite suicide rarely if ever being the main topic [@SymeEtAl2016].

## Study aims and predictions

The primary aim of this study was to examine responses to various forms of emotional signaling in an experimental vignette study. As the costs of a signal are expected to correspond to the reliability of the information provided, we predicted that costlier signals would (1) increase self-reported belief in the signaler's claims and (2) increase the likelihood of participants providing help, with (3) changes in likelihood of support being largely mediated by changes in the participants' belief in the signaler's need from pre-signal to post-signal.

@@

The sole randomized treatment in this study was the signal participants saw, with these signals being one of (1) the control condition: a verbal request without additional signaling, (2) crying, (3) mild-depression, (4) depression, and (5) a suicide attempt, with the order of the conditions here moving from low to high cost.

As many depression scales include measures of crying [for discussion on the relationship between crying and depression see @bylsma2020] and depression is a major risk factor for suicide [@BostwickPankratz2000; @Kessler2012], each signal was designed to build off of the less costly signals so that each signal contained elements of any signal it was costlier than. Although referred to as signals throughout the paper for brevity, we expect that the hypothesized signals of need, like many signals, will also provide information to others in the form of cues [for discussion on the evolution of cues from signals see: @biernaskie2018; @Tiokhin2016; @Steinkopf2015]. The full text for the signals are available in the SI.

@@

To reduce the risk of low quality responses, we used a variety of attention checks, which we clearly labeled in an attempt to avoid demand characteristics on the outcomes of interest. The first was shown immediately after the consent form and provided participants with a random word and asked them to enter the vowels in the order in which they are found in the word. Capitalization did not matter and no word contained repeat vowels. Failure to enter the correct vowels in order resulted in the participants being filtered out of the study. The second followed the vignette and involved asking three questions about the story that should be easy to answer for anyone paying attention.

@@

@@

Responses were recorded with sliders due to the fact they allow for finer grained changes than categorical scales [@KlimekEtAl2017]. Based on findings that a slider's starting position may bias results [@liu2019], we had concerns that participants would be less likely to move away from intermediate starting values than they would be in reality. For this reason, we set each T1 slider to start fully to the left (0). For the exact wording of the question and the labels on the sliders, see table 1.

We also asked which emotions participants felt the signaler was experiencing using a multiple choice question in which they could select as many options as they would like. This included emotions directly related to the signals (e.g., sad, depressed, and suicidal), states which suggest genuine need (traumatized and violated), and states which suggest deception (e.g., deviousness or jealousy). The complete list can be found in the SI.

@@

We asked questions identical to those asked at T1 as the main post-treatment variables of interest (T2). For the *Belief*, *Action*, and *Divide* variables, the position of the slider starting where participants placed it at T1. For the T2 emotion variables this involved no changes from the T1 version.

As both a validity check and a way to understand the degree participants would be willing to help if they believed the participant completely, we presented participants with strong evidence that the claims were true (T3Action). In the US sample, this involved video evidence of the event in question occurring or a similar event after the fact. With the thwarted marriage vignette, this involved the participant seeing the man their daughter wants to marry trash talk their daughter and their family. As a final attention check, we presented participants with the T3Action question but included instructions to ignore this and drag the slider all the way to the left (T4AttentionCheck). We did not exclude anyone from the study for failing this nor did we withhold payment.

@@

The final part of the survey was a brief demographic questionnaire which examined: (1) the age, (2) sex, (3) number of siblings, (4) number of sons, (5) number of daughters, (6) current relationship status, (7) highest level of education, and (8) the annual household income of each participant. The full survey can found in the SI.

@@

The marginal between-subjects effect of the high cost *Suicide attempt* vs. low cost *Verbal request* control on T2 *Belief* (averaging over all four vignettes and all values of T1 *Belief*) was an increase of `r fmt_margins(models$Margins$m1, 'signalSuicide attempt')` points on the original 0-100 point scale. The equivalent increase for T2 *Action* was `r fmt_margins(models$Margins$m3, 'signalSuicide attempt')` points. Given that the standard deviations of T2 *Belief* and *Action* are `r signif(sd(100*d$T2Belief), 2)` and `r signif(sd(100*d$T2Action), 2)`, respectively, these represent increases of `r suicide_belief_eff` and `r suicide_action_eff` SD's, respectively.

@@

Effect sizes (Cohen's d), between- and within-subjects, reveals a similar pattern. The largest between-subjects effect was *Suicide attempt* vs. *Verbal request* control on *Belief* in the "Brother-in-law" vignette, d = `r bil_b_suicide`, and the smallest was the effect of *Crying* on *Belief* in the "Romantic partner" vignette, d =`r rom_b_crying`. The largest within-subjects effect was *Suicide attempt* on *Action* at T2 vs. T1 in the "Brother-in-law" vignette, d = `r bil_a_suicide`, and the smallest was the *negative* effect of *Verbal request* on *Action* at T2 vs. T1 in the "Brother-in-law" vignette, d =`r bil_a_verbal`. See Figures \@ref(fig:effectsizes) and \@ref(fig:effectsizes2).

@@

Consistent with costly signaling theory, costlier signals resulted in greater perceptions of need and increased the likelihood of participants helping the signaler than the less costly signals. This effect was largest in the "brother-in-law" and "romantic partner" vignettes, both of which involved claims of assault against participants' imagined daughters. Compared to these scenarios, the effect of signal type was much smaller in both the "basketball coach" and the "thwarted marriage" vignette. In the "basketball coach" vignette, part of the reason for this may have been the lack of relatedness between participants and the alleged victim. However, this was still a surprising finding due to the relatively small size of the request (i.e., recommend the coach be suspended rather than kick out family members or a romantic partner). With the "thwarted marriage" vignette, the fact that support came at the cost of one's younger daughter may explain the relatively small effect of signal strength. However, data from the Indian sample appears to be of lower quality, limiting our confidence in this interpretation (see limitations for more information). It is also possible that our research in designing the vignettes was insufficient for the population or the content of the vignette.

In the "brother-in-law" and "romantic partner" scenarios, *Crying* had little effect on the magnitude of pro-signaler responses relative to *Verbal request*, suggesting it was not costly enough to serve as a reliable signal in times of substantial conflicts of interests. In contrast, both *Depression* conditions increased support, albeit to similar degrees. One potential reason for this is the increase in symptoms from mild depression to depression was not costly enough to result in a large effect as the symptoms only differed in severity (e.g., grades dropping from As to Bs [mild depression] or Cs [depression]). Such small changes may be less impactful in vignette form than in real-life scenarios in which the effects of signaling may increase in severity and be prolonged over a lengthy period of time.

While the effect of *Suicide attempt* on belief at T2 was similar to the *Depression* conditions across vignettes, it resulted in greater likelihood of support. This may be due to our decision to measure belief in need as the degree participants believed the claims of the signaler (e.g., if the assaults actually happened) rather than a more inclusive measure. For this reason, signal cost may have provided additional information about the costliness of the signaler's adversity on top of information relevant to whether the event claimed actually happened.

Support for this interpretation comes from our mediation analyses, which showed that the likelihood of help was largely, but not entirely, mediated by signal's effect on belief in need in this study but totally mediated by belief in need when measured as need of financial support in the pilot study when comparing *Depression* to *Verbal request*. However, another possibility is that participants viewed the signal as providing information about adversity outside of the signaler's claims. For example, participants who saw the "brother-in-law" vignette may have thought the daughter was lying about the assault due to jealously over her cousin, with the stronger signals still conveying the severity of the competition between her and her cousin.

Overall, our results on the connection between belief in need and likelihood of help are consistent with costly signaling and bargaining approaches to depression and suicide. At T1, *Belief* and *Action* were highly correlated yet distinct; while at T2, the effect of signal on both *Belief* and *Action* was to increase the magnitude of pro-signaler responses, with the effect of signal on *T2Action* being largely mediated by signal's effect on belief in need. This relationship between belief in need and likelihood of support and the importance of signal cost should be expected in scenarios when a target of a signal of need would benefit from helping if the signaler's need was genuine but face substantial net costs if it was not. Importantly, such scenarios appear to be common among modern humans and have likely been since individuals could increase their inclusive fitness through supporting kin or important cooperative partners who could return to providing benefits to supporters once their situation returned to normal [@SugiyamaSugiyama2003; @tooby1996].

>>>>

Major depression and suicide are generally viewed as pathological responses to stress that elicit aversive responses from others. Alternatively, depression and suicide attempts might be the result of adaptations for reliably signaling need and bargaining for increased support from reluctant others. To test this hypothesis, we examined responses to  emotional signals in a vignette study in which the existence of conflicts of interest and private information about one's need were emphasized in four scenarios involving claims of extreme need. Although the effect of signal type varied by vignette, costlier signals like depression and suicide attempts resulted in greater perceptions of need and increased likelihood of support compared to simple verbal requests and crying without further symptoms in our sample of 1,240 participants recruited from Amazon Mechanical Turk. Additionally, the effect of signal type on likelihood of support was largely mediated by the effect of signaling on believe in need as predicted. Our results suggest that depression and suicide attempts, apparent human universals, are a relatively successful way of signaling honest need and eliciting support, consistent with costly signaling explanations.

@@

This strong connection between negative life events and depression along with the adversity and distress that likely accompanies these events has led to the suggestion that this relationship is causal [@KendlerEtAl1999]. Support for this comes from the fact that the relationship between negative events and depression holds true when looking solely at events outside of one's control, showing that the connection is unlikely to be driven solely by depressed individuals creating situations where negative events are likely to be common [@Hammen2005; @KendlerEtAl1999]. In addition to this, twin studies have shown that one's history of negative events is still a strong predictor of major depression when controlling for genetic similarity and that part of the heritability of depression stems from the heritability of negative events, like divorce and family conflict [@KendlerBaker2007; @KendlerEtAl1999].

Of the types of negative life events associated with depression, those which involve conflict tend to be more strongly associated with depression [@Hammen2005; @Mazure1998]. Marital problems, bullying, and the existence of abusive relationships are all commonly cited risk factors for depression [@KendlerEtAl1995; @KendlerEtAl1999; @klomek2007; @klomek2019], with  assault (both sexual and non-sexual), in particular, greatly increasing one's risk of depression [@KendlerEtAl1995; @KendlerEtAl1999].  Other notable risk factors like loss of a loved one, financial problems, job loss, or severe or prolonged illness, might seem less related to conflict. However, the ability for others to provide meaningful support to limit the harm caused in times of need [@SugiyamaSugiyama2003] combined with the ubiquity of conflicts of interests mean that many types of problems will eventually involve social conflict even if they did not start that way [@Hagen2003)]. In line with this reasoning, conflict, in general, was found to be associated with depression among the Tsimane, particularly when it involved non-kin [@StieglitzEtAl2015a].

@@

One possible reason for the connection between depression, suicide, and social adversity is that some instances of depression and suicide attempts serve as means to elicit increased support from others [@andrews2006; @Hagen2002; @Hagen2003; @HagenEtAl2008; @nock2008; @Rosenthal1993; @Stengel1956; @SymeEtAl2016], particularly in times when an individual cannot solve substantial fitness-relevant challenges on their own [@Hagen2003; @SymeEtAl2016]. According to the bargaining models of depression and suicide, the idea is that the small groups humans lived in for much of our evolutionary history created many situations in which certain forms of cooperation were unlikely to be easily replaced by others while also allowing for harmful or unfair social relationships to develop [@Hagen2003; @SymeEtAl2016]. In these instances, one's monopoly on one or more types of cooperation could be leveraged to change how one is treated if others valued an individual's cooperation sufficiently and believed them to be in genuine need.

Due to the ease in which signals can be faked [@Zahavi1975; @mokkonen2016] and the costs of accepting a deceptive signal as reliable [@dawkins1978; @johnstone1993; @Hasson1994], demonstrating that one’s need is genuine has likely been a recurrent adaptive problem since humans evolved the capacity to provide instrumental support to one another. One possible solution to this is signaling in ways that are unprofitable to fake due to the costs involved. [@Zahavi1975]. This is possible whenever the signal in question is comparatively cheaper to produce for those who have the underlying quality being compared to those who lack it [@grafen1990]. In these instances, individuals with the quality maximize their fitness at greater levels of investment in the signal compared to those who lack the quality or possess it to a lesser degree [@zahavi1977]. The end result of this is expected to be a signal that either tends to not be worth faking or is impossible to fake to a large degree due to selection against fakers constraining the physiology behind the signal [@zahavi1977; @gangestad2005].

Both depression and suicidal behavior fit this criterion as they are both relatively costly for those not facing adversity than those who are. In the case of depression, the largescale reduction in one's normal routine of fitness enhancing behavior is relatively less costly for those whose situations result in a limited ability to obtain fitness-relevant benefits, with depression likely adding no additional costs in instances of extreme deprivation [@Hagen2003]. When it comes to suicide, the bulk of the costs associated with suicide attempts are only felt in times that result in death or severe bodily harm. For this reason, the way others view and respond to suicide attempts is likely to be strongly influenced by the probability that the attempt would result in death or severe injury [@Rosenthal1993]. Despite this difference, the previously mentioned logic still applies with increased risk of death being relatively more costly for those able to advance their fitness prospects in their day-to-day lives.

The costs of depression and suicidal behavior may be particularly important in times of conflict or when the level of support needed to address one’s situation is substantial, both of which should be expected to increase the payoff of manipulating others through fake signals. When either condition applies, less costly signals like verbal requests, facial sadness, and crying are likely to be less reliable as their low costs can easily be made up for with increased social support, thereby favoring the use of costlier signals, so long as the costs of signaling are outweighed by the benefits [@SymeHagen2021].

Beyond signaling need to those invested in the well-being of the signaler, depression and suicidal behavior may also help an individual bargain for increased support [@Hagen2003]. This is possible due to the reduced investment in others that frequently accompanies depression or the threat to future cooperation that comes from a suicide. In these instances, depression and suicide attempts are suggested to function analogously to a labor strike, sending the message that one’s need is genuine and that if the target of the signal wanted to continue receiving benefits from the depressed, he or she would need to increase support towards the signaler.

@@

Distinguishing between the information provided by signals versus cues is a major challenge for signaling approaches to behavior. A signal differs from a cue in that it is intentionally sent by an organism to alter the behavior or thoughts of other individuals, while a cue is any piece of information an organism uses to alter its behavior or physiological state that was created without the intent to do so [@MaynardSmithHarper1995]. Regarding the signals of interest, suicide attempts are more likely to take the form of pure, external signals than other symptoms of depression. This is because low mood's impact on behavior for non-signaling benefits likely conveys information as cues in addition to what might exist for signaling purposes. In other words, the signaling component of depression may come from the upregulation of symptoms beyond what would be optimal for non-signaling purposes rather than the creation of entirely new symptoms. Suicidality is likely an exception to this, due to the likelihood most suicide attempts likely do not have non-signaling functions (see @deCatanzaro1991 for a possible exception).

@@

For depression and suicide attempts to be worth the costs involved with signal creation, meaningful social support would be required, either in the form of temporary but impactful support or long-term changes in how one is treated [@Hagen2003; @SymeEtAl2016]. Most studies examining responses to depression have been observational and focused on responses outside of changes in social support. In general, most report aversive responses and high levels of rejection towards the depressed, and depressed individuals are often linked to impaired social functioning [@EvraireDozois2011; @GotlibLee1989; @HirschfeldEtAl2000]. One explanation for this is that depressed individuals engage in behaviors that drive others away. In line with this reasoning, both excessive reassurance seeking and negative feedback seeking, which often accompany depression, have been shown to elicit negative emotional responses from others and increase the chance of rejection from others [@JoinerMetalsky1995; @StarrDavila2008]. Negative interactions are also seen in families as depressed children tend to report less support and more negative responses from parents than non-depressed children, although it has not been demonstrated that such interactions did not lead to depression in the first place [@MesserGross1995; @SchwartzEtAl2012].

Despite being looked for less frequently, beneficial responses to depressed individuals are also reported. This is seen in that depression has been associated with reduced likelihood of aggression within families [@hops1987; @dadds1992], including those already experiencing high levels of conflict [@SheeberEtAl2001]. There is also limited evidence that depression can result in increased levels of support from others outside the family. For example, non-depressed individuals were shown to be more cooperative and exhibit more caretaking behavior towards their college depressed roommates over the course of their time living together [@HokansonEtAl1986]. Likewise, a study examining responses to confederates trained to appear ‘depressed’, ‘anxious’, or ‘normal’ found that individuals were more likely to provide advice and support to ‘depressed’ individuals than ‘anxious’ or ‘normal’ individuals despite the confederate in the ‘depression’ condition being viewed more negatively than the other conditions and the ‘depression’ condition involving similar levels of distress compared to the ‘anxious’ condition [@stephens1987].

Most studies examining responses to those who have survived suicide attempts have focused on stigmatization. In these studies, perceptions of survivors as being weak, selfish, mentally ill, and antisocial are all being commonly reported [@BatterhamEtAl2013; @TzengLipson2004], with stigmatization often being found within and outside one's social network [@FreyEtAl2016; @ScoccoEtAl2012]. Despite this potential for stigmatization, increased social support and changes to important relationships have been reported to follow suicide attempts with some indication that these effects may hold long term [@Stengel1956]. For example, a study of 100 women who survived suicide attempts found that individuals gained identifiable benefits through the attempt in 75 cases, with 41 individuals benefiting from reconciliations with others [@Lukianowicz1971]. Unlike many Western countries, where suicide is often viewed as pathological [@Hidaka2012], members of traditional societies have been reported to view suicide attempts as cries for help rather than mental illness [@Shostak1981], with attempts having been described as ways of escaping unwanted marriage arrangements, persistent abuse, or a lack of support in obtaining mates by those engaging in the behavior and observers [@GutierrezdePinedaMuirden1948; @Hilger1957; @Karsten1935; @Tessmann1930; @Wilson1960]. Although findings that individuals view suicide attempts as cries for help is not necessarily evidence for them resulting in beneficial responses, 30 out of 84 examples of suicidal behavior included in the HRAF resulted in positive changes for the survivor, even though suicide was rarely if ever the main topic [@SymeEtAl2016].

# Study aims and predictions

The primary aim of this study was to examine responses to various forms of emotional signaling in an experimental vignette study. We predicted that costlier signals would increase the participants' self-reported belief in the signaler's claims and the likelihood of them providing help, with this prediction entailing the following rank from low to high cost: verbal request, crying, mild-depression, depression, and suicide attempt. We also predicted that changes in the likelihood participants would provide support would be mediated by changes in the participants' belief in the signaler's need pre-signal to post-signal.

@@

The sole randomized treatment in this study was the signal participants saw, with these signals being one of (1) the control condition: a verbal request without additional signaling, (2) crying, (3) mild-depression, (4) depression, and (5) a suicide attempt.

As many depression scales include measures of crying [for discussion on the relationship between crying and depression see @bylsma2020] and depression is a major risk factor for suicide [@BostwickPankratz2000; @Kessler2012], each signal was designed to build off of the less costly signals so that each signal contained elements of any signal it was costlier than. Although referred to as signals throughout the paper for brevity, we expect that the hypothesized signals of need, like many signals, will also provide information to others in the form of cues [For discussion on the evolution of cues from signals see: @biernaskie2018; @Tiokhin2016; @Steinkopf2015]. The full text for the signals are available in the SI.

@@

To reduce the risk of low quality responses, we used a variety of attention checks, which we clearly labeled in an attempt to avoid demand characteristics. The first was shown immediately after the consent form and provided participants with a random word and asked them to enter the vowels in the order in which they are found in the word. Capitalization did not matter and no word contained repeat vowels. Failure to enter the correct vowels in order resulted in the participants being filtered out of the study. The second followed the vignette and involved asking three questions about the story that should be easy to answer for anyone paying attention.

@@

After accepting the HIT on MTurk, every participant took the survey using Qualtrics. In every instance, the elements of the survey were laid out in the following order: 1) the consent form, 2) the vowel attention check, 3) the main body of the vignette, 3) the pre-test baseline measures (T1), 4) the story content attention check, 5)  the signal portion of the vignette, 7) the post-test outcome measures (T2-T3), 8) an attention check that did not filter out participants (T4), 9) a set of demographic questions, and 10) an open feedback prompt. The full survey can be found in the SI.

@@

Responses were recorded with sliders due to the fact they allow for finer grained changes than categorical scales [@KlimekEtAl2017]. Due to findings that a slider's starting position may bias results [@liu2019], each T1 slider started fully to the left (0) (don't exactly remember the logic for left instead of click to start). For the exact wording of the question and the labels on the sliders, see table 1.

We also asked which emotions participants felt the signaler was experiencing using a multiple choice question in which they could select multiple options. This included emotions directly related to the signals (e.g., sad, depressed, and suicidal), states which suggest genuine need (traumatized and violated), and states which suggest deception (e.g., deviousness or jealousy). The complete list can be found in the SI.

@@

We asked questions identical to those asked at T1 as the main post-treatment variables of interest (T2). For the belief, action, and divide variables, the position of the slider starting where participants placed it at T1. For the T2 emotion variables this involved no changes from the T1 version.

As both a validity check and a way to understand the degree participants would be willing to help if they believed the participant completely, we presented participants with strong evidence that the claims were true (T3Action). In the US sample, this involved video evidence of the event in question occurring or a similar event after the fact. With the thwarted marriage vignette, this involved the participant seeing the man their daughter wants to marry trash talk their daughter and their family. As a final attention check, we presented participants with the T3Action question but included instructions to ignore this and drag the slider all the way to the left (T4AttentionCheck). We did not exclude anyone from the study for failing this nor did we withhold payment due to the time participants had invested in the study at this point.

@@

The final part of the survey was a brief demographic questionnaire which examined: (1) the age, (2) sex, (3) number of siblings, (4) number of sons, (5) number of daughters, (6) current relationship status, (7) highest level of education, and (8) the annual household income of each participant.

@@

The marginal between-subjects effect of the high cost *Suicide attempt* vs. the *Verbal request* control on T2 *Belief* (averaging over all four vignettes and all values of T1 *Belief*) was an increase of `r fmt_margins(models$Margins$m1, 'signalSuicide attempt')` points on the original 0-100 point scale. The equivalent increase for T2 *Action* was `r fmt_margins(models$Margins$m3, 'signalSuicide attempt')` points. Given that the standard deviations of T2 *Belief* and *Action* are `r signif(sd(100*d$T2Belief), 2)` and `r signif(sd(100*d$T2Action), 2)`, respectively, these represent increases of `r suicide_belief_eff` and `r suicide_action_eff` SD's, respectively.

@@

The largest between-subjects effect was *Suicide attempt* vs. *Verbal request* control on *Belief* in the "Brother-in-law" vignette, d = `r bil_b_suicide`, and the smallest was the effect of *Crying* on *Belief* in the "Romantic partner" vignette, d =`r rom_b_crying`. The largest within-subjects effect was *Suicide attempt* on *Action* at T2 vs. T1 in the "Brother-in-law" vignette, d = `r bil_a_suicide`, and the smallest was the *negative* effect of *Verbal request* on *Action* at T2 vs. T1 in the "Brother-in-law" vignette, d =`r bil_a_verbal`. See Figures \@ref(fig:effectsizes) and \@ref(fig:effectsizes2).

@@

Consistent with costly signaling theory, costlier signals resulted in greater perceptions of need and increased the likelihood of participants helping the signaler than the less costly signals. This effect was largest in the "brother-in-law" and "romantic partner" vignettes, both of which involved claims of assault against participants' imagined daughters. However, the effect of signal type was much smaller in both the "basketball coach" and the "thwarted marriage" vignette. In the "basketball coach" vignette, part of the reason for this may have been the lack of relatedness between participants and the alleged victim. However, this was still a surprising finding due to the relatively small size of the request (i.e., recommend the coach be suspended rather than kick out family members or a romantic partner). With the thwarted marriage vignette, the fact that support came at the cost of one's younger daughter may explain the relatively small effect of signal strength. However, data from the Indian sample appears to be of lower quality, limiting our confidence in this interpretation (see limitations for more information). It is also possible that our research in designing the vignettes was insufficient for the population or the content of the vignette.

In the "brother-in-law" and "romantic partner" scenarios, *Crying* had little effect on the magnitude of pro-signaler responses, while both *Depression* conditions increased support to similar degrees. One potential reason for this is the increase in symptoms from mild depression to depression was not costly enough to result in a large effect as the symptoms only differed in severity (e.g., grades dropping from As to Bs [mild depression] or Cs [depression]). Such small changes may be less impactful in vignette form than in real-life scenarios in which the effects of signaling may increase in severity and be prolonged over a lengthy period of time.

While the effect of *Suicide attempt* on belief at T2 was similar to the *Depression* conditions, it resulted in greater likelihood of support. This may be due to our decision to measure belief in need as the degree participants believed the claims of the signaler (i.e., the assaults or incompatibility with a potential husband in the context of arranged marriage) rather than a more inclusive measure. For this reason, signal cost may have provided additional information about the costliness of the signaler's adversity in ways that would still be consistent with signaling theory.

Support for this interpretation comes from our mediation analyses, which showed that the likelihood of help was largely, but not entirely, mediated by signal's effect on belief in need in this study but totally mediated by belief in need when measured as belief in need of financial support in the pilot study. However, the incomplete mediation in this study may have arisen if participants viewed the signal as providing information about adveristy outside of the signaler's claims. For example, participants who saw the "brother-in-law" vignette may have thought the daughter was lying about the assault due to jealously over her cousin, with the stronger signals still conveying the severity of the competition between her and her cousin.

## Broader discussion

Our results on the connection between belief in need and likelihood of help are consistent with costly signaling and bargaining approaches to depression and suicide. In addition to the effect of signal cost on support being largely mediated by the signal's effect on belief in need, belief in need and likelihood of support were highly correlated but distinct at T1. This situation should arise in any instance when a target of a signal of need would benefit from helping if the need was genuine and known to reasonable precision. Importantly, such problems are common among modern humans and have likely been so since individuals could increase their inclusive fitness through supporting kin or important cooperative partners in times of genuine need [@SugiyamaSugiyama2003; @tooby1996].
